A new preprocedural predictive risk model for post-endoscopic retrograde cholangiopancreatography pancreatitis: The SuPER model

Version 1

Abstract

Abstract

Background

Post–endoscopic retrograde cholangiopancreatography (ERCP) pancreatitis (PEP) is a severe and deadly adverse event following ERCP. The ideal method for predicting PEP risk before ERCP has yet to be identified. We aimed to establish a simple PEP risk score model (SuPER model: Support for PEP Reduction) that can be applied before ERCP.

Methods

This multicenter study enrolled 2074 patients who underwent ERCP. Among them, 1037 patients each were randomly assigned to the development and validation cohorts. In the development cohort, the risk score model for predicting PEP was established by logistic regression analysis. In the validation cohort, the performance of the model was assessed.

Results

In the development cohort, five PEP risk factors that could be identified before ERCP were extracted and assigned weights according to their respective regression coefficients: -2 points for pancreatic calcification, 1 point for female sex, and 2 points for intraductal papillary mucinous neoplasm, a native papilla of Vater, or the use of pancreatic duct procedures. The PEP occurrence rate was 0% among low-risk patients (≤ 0 points), 5.5% among moderate-risk patients (1 to 3 points), and 20.2% among high-risk patients (4 to 7 points). In the validation cohort, the C-statistic of the risk score model was 0.71 (95% CI 0.64-0.78), which was considered acceptable. The PEP risk classification (low, moderate, and high) was a significant predictive factor for PEP that was independent from intraprocedural PEP risk factors (precut sphincterotomy and inadvertent pancreatic duct cannulation) (OR 4.2, 95% CI 2.8-6.3, P < 0.01).

Conclusions

The PEP risk score allows an estimation of the risk of PEP prior to ERCP, regardless of whether the patient has undergone pancreatic duct procedures. This simple risk model, consisting of only five items, may aid in predicting and explaining the risk of PEP before ERCP and in preventing PEP by allowing selection of the appropriate expert endoscopist and useful PEP prophylaxes.

Introduction

Endoscopic retrograde cholangiopancreatography (ERCP) is widely performed as an important diagnostic and therapeutic procedure for pancreaticobiliary diseases. ERCP-related procedures are relatively risky among endoscopic procedures. The high-risk adverse events of ERCP include duodenal perforation and bleeding after endoscopic sphincterotomy (EST) and post-ERCP pancreatitis (PEP). The rate of PEP occurrence is 3.1-13.0% (Andriulli et al., 2007, Freeman et al., 1996, Glomsaker et al., 2013, Katsinelos et al., 2014, Kochar et al., 2015, Loperfido et al., 1998). PEP can even become life-threatening. The fatality rate of PEP is 0.1–0.7% (Andriulli et al., 2007, Kochar et al., 2015). Therefore, the decision to perform ERCP should be made carefully, considering each patient’s risk factors for PEP.

To predict an individual patient’s PEP risk, five scoring systems have been devised (Chiba et al., 2021, DiMagno et al., 2013, Friedland et al., 2002, Fujita et al., 2021, Zheng et al., 2020). The first risk scoring system for PEP occurrence was established in 2002. In that study, pain during the procedure, pancreatic duct cannulation, a history of PEP, and the number of cannulation attempts were identified as risk factors for PEP. After the first scoring system was reported, each new scoring system used risk factors that were extracted by multivariate analyses. These included various patient characteristics before ERCP and postprocedural risk factors. As postprocedural risk factors, precut sphincterotomy and difficult cannulation were proposed, but it is difficult to predict these risk factors and to determine the PEP risk before ERCP. Thus, a new prediction scoring system for PEP before ERCP is desirable. If the risk of PEP could be predicted before ERCP, then the expert endoscopist can perform ERCP from the start, and high-PEP-risk procedures (for example, precut sphincterotomy, multiple cannulation attempts, inadvertent pancreatic duct cannulation) can be avoided (Testoni et al., 2010, Wang et al., 2009). If biliary cannulation without the use of at least one high-PEP-risk procedure is difficult, other treatments (for example, percutaneous transhepatic biliary drainage (PTBD) or endoscopic ultrasound (EUS)-guided biliary drainage (EUS-BD)) could be considered.

Therefore, we aimed to establish a PEP prediction model using only risk factors that can be gathered before ERCP. Our model was developed and validated with multicenter data from Japan.

Methods

We performed a multicenter retrospective study at six institutions in Japan. This study was approved by the institutional review board of Fukushima Medical University and that of each partner medical institution. All patients agreed to undergo ERCP after providing written consent.

Patients

Among 2,176 patients who underwent ERCP between November 2020 and October 2022, 2,074 were enrolled in this study. The other 102 patients were excluded for the following reasons: past history of choledochojejunostomy, acute pancreatitis, choledochoduodenal fistula, difficulty finding the Vater papilla, past history of pancreatojejunostomy, or past history of pancreatogastrostomy (Figure 1).

Flowchart of the inclusion criteria. ERCP, endoscopic retrograde cholangiopancreatography.

Study design

We randomly sampled 50% of the patients as the development cohort and 50% as the validation cohort (Figure 1). In the development cohort, we established a risk scoring system for predicting PEP before ERCP, which was named the support for PEP reduction model (SuPER model). The validation cohort was used to confirm the effectiveness of the scoring system. PEP diagnosis and severity were assessed according to Cotton’s criteria (Cotton et al., 1991). Patients who experienced abdominal pain and had hyperamylasemia (more than three times the normal upper limit) at least 24 hours after ERCP were diagnosed with PEP. Mild PEP was defined as pancreatitis that required prolongation of the planned hospitalization by 2-3 days. Moderate PEP was defined as pancreatitis that required 4-10 days of hospitalization. Severe PEP was defined as pancreatitis that required more than 10 days of hospitalization or intervention or hemorrhagic pancreatitis, phlegmon, or pseudocysts.

To establish the risk score, the risk factors for PEP were investigated using the data from the development cohort. To determine the PEP risk score, factors that might be associated with PEP occurrence were investigated. To predict the PEP risk score before ERCP, factors related to patient characteristics and previously scheduled procedures, as reported in past studies, were selected. The patients’ risk factors included age < 50 years, female sex, a past history of pancreatitis, a past history of PEP, a past history of gastrectomy, pancreatic cancer, intraductal papillary mucinous neoplasm (IPMN), a native papilla of Vater, absence of chronic pancreatitis (CP), normal serum bilirubin (≤ 1.2 mg/dl), and periampullary diverticulum (Ding et al., 2015, Freeman et al., 2001, Freeman et al., 1996, Fujita et al., 2022, Fujita et al., 2021, Masci et al., 2003, Wang et al., 2009, Williams et al., 2007, Zheng et al., 2020). Pancreatic divisum was excluded from the patient risk factor list because pancreatic divisum was observed in only two patients. Pancreatic calcification and a diameter of the main pancreatic duct > 3 mm were considered to indicate CP (Beyer et al., 2023, Sarner and Cotton, 1984).

These imaging findings were confirmed by CT, MRI, or EUS before ERCP. The CT and MRI findings were reviewed by radiologists. IPMN was diagnosed according to the results of CT, MRI, and EUS. As pre-ERCP prophylaxes for PEP, gabexate or nafamostat, intravenous hydration, and NSAID suppositories were used (Fujita et al., 2022). As planned procedure-related risk factors, EST, endoscopic papillary balloon dilation (EPBD), endoscopic papillary large balloon dilation (EPLBD) using a ≥ 12 mm balloon catheter (Itoi et al., 2018), biliary stone removal, ampullectomy, biliary stent material (plastic stent, self-expandable metallic stent (SEMS), or covered SEMS (CSEMS)), inside stent placement, and procedures on the pancreatic duct were evaluated (Freeman et al., 2001, Freeman et al., 1996, Harewood et al., 2005, Kato et al., 2022, Masci et al., 2003, Masci et al., 2001, Testoni et al., 2010, Williams et al., 2007). A biliary stent above the Vater papilla was also assessed as a prophylactic measure against PEP (Ishiwatari et al., 2013).

To demonstrate the independence of the established risk classification, the relationship between it and intraprocedural PEP risk factors (including precut sphincterotomy and inadvertent pancreatic duct cannulation) (Testoni et al., 2010, Wang et al., 2009) were investigated. Because of the retrospective nature of the data, the exact cannulation times and the number of cannulation attempts were not available.

Therefore, multiple cannulation attempts and a prolonged cannulation time could not be investigated as intraprocedural PEP risk factors.

Sample size

The primary aim of this study was to establish a PEP prediction model that could be used to calculate a risk score before ERCP. To construct a prediction model by logistic regression analysis, 10 events per explanatory variable were needed (Wynants et al., 2015). Seven variables were evaluated in the development cohort, so 70 PEP patients were required. Five variables were evaluated in the validation cohort, so 50 PEP patients were necessary for it. According to a previous systematic review, the rate of PEP occurrence was 9.7% (Kochar et al., 2015). Therefore, at least 722 and 521 patients were included in the development and validation cohorts, respectively.

Statistical analysis

In the development cohort, univariate and multivariate logistic regression analyses were performed to identify the risk factors for PEP. The factors that had a p value < 0.10 in the univariate analysis were included in the multivariate analysis. To construct the scoring system for PEP risk, the factors with p < 0.10 in the multivariate analysis were ultimately included in the risk score model. The factors selected in the multivariate analysis were assigned points according to the regression coefficient (each variable’s risk points = the ratio of the variable’s regression coefficient/minimum regression coefficient). The sum of the assigned points was calculated for each patient, and the patients were classified into three groups (low risk, moderate risk, and high risk) according to the expected rate of PEP occurrence (Friedland et al., 2002). The risk classification system (SuPER model) was also applied to the validation cohort.

With respect to both the development and validation cohorts, the effectiveness of the risk score model was evaluated as follows. The correlations between the risk score, risk classification and PEP occurrence were evaluated by the Cochran–Armitage trend test. The predictive accuracy of the risk score was assessed using the C statistic. The goodness of fit of the model was evaluated using the Hosmer‒Lemeshow test. The independence of the established risk classification from the unexpected intraprocedural PEP risk factors was assessed by multivariate logistic regression analyses.

Patients with missing data for variables selected in the risk score model were removed from the final cohort.

Statistical analyses were performed using EZR version 1.62 (Saitama Medical Centre, Jichi Medical University, Saitama, Japan) and SPSS version 26.0 (IBM Corp., Armonk, NY, USA). A p value < 0.05 indicated statistical significance.

Results

Patient characteristics and ERCP outcomes in each cohort

The patient characteristics and ERCP outcomes in each cohort are shown in Table 1. A total of 1037 patients were assigned to each of the development and validation cohorts, including 70 (6.8%) and 64 (6.2%) patients diagnosed with PEP, respectively. The pre-ERCP prophylactic measures used at each hospital differed, and not all patients received prophylaxis.

Construction of the PEP risk scoring system

According to the univariate analyses, age < 50 years, female sex, IPMN, a native papilla of Vater, pancreatic calcification, EST, and procedures on the pancreatic duct had p values < 0.10 (Table 2). According to the multivariate analysis, female sex, IPMN, a native papilla of Vater, pancreatic calcification, and procedures on the pancreatic duct had p values < 0.10. These factors were assigned risk points according to their respective regression coefficients.

The risk score of each patient was calculated as the total of that patient’s risk points and ranged from -2 to 7 points (Table 3). The risk score was found to be correlated with PEP occurrence (p < 0.01, Cochran–Armitage trend test). The patients were classified as low (≤ 0 points), moderate (1-3 points), or high risk (4-7 points) for PEP according to the risk score. The PEP rates were 0% (0/327) among the low-risk patients, 5.5% (27/492) among the moderate-risk patients, and 20.2% (39/193) among the high-risk patients. The risk classification was correlated with PEP occurrence (p < 0.01, Cochran–Armitage trend test).

The C statistic of the risk score model was sufficiently high at 0.77 (95% confidence interval (CI) 0.72-0.82) (Table 4). The goodness of fit of the risk score model was also confirmed by the Hosmer–Lemeshow test (p = 0.59).

Validation of the PEP risk scoring system

The risk score was associated with PEP occurrence in the validation cohort (p < 0.01, Cochran–Armitage trend test) (Table 3). We found that 2.4% (8/331) of the patients at low risk, 5.3% (27/513) of those at moderate risk, and 18.0% (29/161) of those at high risk experienced PEP. The risk classification was also correlated with PEP occurrence in the validation cohort (p < 0.01, Cochran–Armitage trend test).

The C statistic of the risk score was 0.71, which was also high in the validation cohort (Table 4). The PEP risk score model showed good fitness according to the Hosmer–Lemeshow test (p = 0.40). According to the above results, the preprocedural PEP risk could be calculated, as shown in Figure 2.

Example of the preprocedural PEP risk checklist. ERCP, endoscopic retrograde cholangiopancreatography. IPMN, intraductal papillary mucinous neoplasm; PEP, post-ERCP pancreatitis.

Risk classification and unexpected PEP risk factors

The relation between the established risk classification and intraprocedural PEP risk factors is shown in Appendix 1—table 1. In all patients, the development cohort, and the validation cohort, the risk classification was significantly associated with the occurrence of PEP. On the other hand, precut sphincterotomy and inadvertent pancreatic duct cannulation were not significantly associated with the occurrence of PEP.

Discussion

In this multicenter study, we created a risk scoring system (the SuPER model) using five items that could be measured before performing ERCP. With this score, PEP occurrence could be accurately predicted to some degree. Besides, the established PEP risk classification was associated with PEP occurrence independently from unpredictable intraprocedural PEP risk procedures.

This risk scoring and classification of PEP has several advantages. First, the score is calculated using only five items, all of which can be easily assessed by medical interviews and imaging (for example, CT). One scoring system included sphincter of Oddi dysfunction (SOD) as a test item (DiMagno et al., 2013). The diagnosis of SOD requires sphincter of Oddi manometry and fulfilment of the criteria for biliary pain, but sphincter of Oddi manometry is not widely used (Cotton et al., 2016). The diagnostic criterion for biliary pain included 8 items, and that for SOD included 15 items. Among the items of the SuPER risk scoring system, pancreatic calcification was assigned -2 points. Its low weighting could be explained by the following. The international conceptual model of CP can be divided into four stages: acute pancreatitis–recurrent acute pancreatitis, early CP, established CP, and end-stage CP (Whitcomb et al., 2016). Established CP patients have already passed the acute pancreatitis–recurrent acute pancreatitis course, and pancreatic calcification has been reported in established CP patients. Acinar dysfunction has also been observed in these patients (Whitcomb et al., 2016). Therefore, patients with pancreatic calcification may have a lower incidence of PEP.

Second, the SuPER risk score can be determined before the ERCP procedure, as the established risk classification was found to be the sole significant factor predicting the occurrence of PEP independent from intraprocedural PEP risk factors. As described in the Background section, precut sphincterotomy, multiple cannulation attempts, and a cannulation time greater than 10 minutes were identified as high risk factors that cannot be accounted for prior to ERCP (Testoni et al., 2010, Wang et al., 2009). Although the established PEP risk classification was independent from the included intraprocedural risk factors (precut sphincterotomy and inadvertent pancreatic duct cannulation), detailed data on the number of cannulation attempts and the cannulation time were not available. Therefore, to avoid intraoperative procedures associated with a high risk of PEP occurrence, an expert endoscopist can initially perform ERCP for high-PEP-risk patients. In addition, PEP prophylaxis can be administered beforehand for high-PEP-risk patients. As effective prophylaxes for PEP, rectal NSAIDS and pancreatic stent placement have been reported (Elmunzer et al., 2008, Murray et al., 2003, Sugimoto et al., 2019). In this report, rectal NSAID use was not identified as a significant factor preventing PEP. One reason for this is that in past reports describing the use of rectal NSAIDs to prevent PEP, patients at high risk for PEP were often treated (Elmunzer et al., 2008). In contrast, this study included all patients who underwent ERCP. Another reason might be the difference in dose. One hundred milligrams of rectal diclofenac was used in past reports, whereas 12.5-50 mg of diclofenac was used in this study. In Japan, the approved diclofenac dose covered by insurance is 50 mg or less, with the dose being typically lower for elderly patients.

Therefore, diclofenac doses of 12.5-50 mg were prescribed by the doctors depending on the age and size of the patients. Pancreatic stent placement itself is one of the procedures performed on the pancreatic duct and was a higher-risk procedure for PEP than endoscopic biliary procedures without an approach to the pancreatic duct (Appendix 2—table 2). Moreover, pancreatic stent placement has become a prophylactic treatment for PEP in patients who have undergone pancreatography or wire placement to the pancreatic duct (Mazaki et al., 2014, Sugimoto et al., 2019). As described above, pancreatic stent placement was performed along with high-risk-PEP procedures (i.e., guidewire placement to the pancreatic duct or pancreatography); therefore, pancreatic stent placement was grouped together with the other endoscopic retrograde pancreatography procedures as “procedures on the pancreatic duct”.

This study has several limitations. First, the study was retrospective, and there were missing data. However, the results reported are trustworthy. The percentage of patients who did not meet the inclusion criteria was not more than 5%, and the percentage of missing data was not over 1%. As described in the Materials and Methods section, patients with missing data for the variables selected in the risk score model were removed from the final cohort. The reliability of the SuPER risk score model was also statistically confirmed. Second, some factors cannot be assessed before ERCP. Additional procedures could be conducted during ERCP, and unplanned pancreatography is often performed in patients who are scheduled for endoscopic cholangiography or biliary treatment. However, the established PEP risk classification was independent from the included intraprocedural risk factors. A planned procedure for accessing the pancreatic duct is listed in the SuPER risk model. Therefore, we can predict the SuPER risk score and classification of patients regardless of whether they have undergone pancreatic duct procedures. Third, this study was performed in a single country. Validation studies over wider geographic regions are necessary.

In conclusion, a simple and useful PEP scoring system (SuPER model) with only five clinical items was developed in this multicenter study. This scoring system may aid in predicting and explaining PEP risk and in selecting appropriate prophylaxes for PEP and endoscopic pancreaticobiliary procedures for each patient.

eLife Assessment

This valuable study discusses a hot topic in post-endoscopic retrograde cholangiopancreatography pancreatitis. The new score for predicting post-ERCP pancreatitis offers an idea about the risk of pancreatitis before the procedure. Although most scores depend on intraprocedural manoeuvres, such as the number of attempts to cannulate the papilla, this is a solid retrospective single-center study in one country. To be validated, this score should be done in many countries and on large numbers of patients, nevertheless, this paper should interest gastrointestinal endoscopists.

Joint Public Review:

Summary:

This work provides a new general tool for predicting post-ERCP pancreatitis before the procedure depending on pancreatic calcification, female sex, intraductal papillary mucinous neoplasm, a native papilla of Vater, or the use of pancreatic duct procedures. Even though it is difficult for the endoscopist to predict before the procedure which case might have post-ERCP pancreatitis, this new model score can help with the maneuver and when the patient is at high risk of pancreatitis, sometimes can be deadly), so experienced endoscopists can do the procedure from the start. This paper provides a model for stratifying patients before the ERCP procedure into low, moderate, and high risk for pancreatitis. To be validated, this score should be done in many countries and on large numbers of patients. Risk factors can also be identified and added to the score to increase rank.

Strengths:

(1) One of the severe complications of endoscopic retrograde cholangiopancreatography procedure is pancreatitis, so investigators try all the time to find a score that can predict which patients will probably have pancreatitis after the procedure. Most scores depend on the intraprocedural maneuver. Some studies discuss the preprocedural score that can predict pancreatitis before the procure. This study discusses a new preprocedural score for post-ERCP pancreatitis.

(2) Depending on this score that identifies low, moderate, and high-risk patients for post-pancreatitis, so from the start, experienced and well-trained endoscopists can do the procedure or can refer patients to tertiary hospitals or use interventional radiology or endoscopic retrograde cholangiopancreatography.

(3) The number of patients in this study is sufficient to analyze data correctly.

Weaknesses:

(1) It is a single-country, retrospective study.

(2) Many cases were excluded, so the score cannot be applied to those patients.

(3) Many other studies, e.g., https://link.springer.com/article/10.1007/s00464-021-08491-1, https://pubmed.ncbi.nlm.nih.gov/36344369/, that have been published before discussing the same issue, so what is the new with this score?

(4) The discussion section needs reformulation to express the study's aim and results.

(5) Why did the authors select these items in their scoring system and did not add more variables?

Author response:

The following is the authors’ response to the original reviews.

Joint Public review:

Summary:

This work provides a new general tool for predicting post-ERCP pancreatitis before the procedure depending on pancreatic calcification, female sex, intraductal papillary mucinous neoplasm, a native papilla of Vater, or the use of pancreatic duct procedures. Even though it is difficult for the endoscopist to predict before the procedure which case might have post-ERCP pancreatitis, this new model score can help with the maneuver and when the patient is at high risk of pancreatitis, sometimes can be deadly), so experienced endoscopists can do the procedure from the start. This paper provides a model for stratifying patients before the ERCP procedure into low, moderate, and high risk for pancreatitis. To be validated, this score should be done in many countries and on large numbers of patients. Risk factors can also be identified and added to the score to increase rank.

Thank you for reviewing our manuscript. We hope that this score will be validated in other countries from now on.

Strengths

(1) One of the severe complications of endoscopic retrograde cholangiopancreatography procedure is pancreatitis, so investigators try all the time to find a score that can predict which patients will probably have pancreatitis after the procedure. Most scores depend on the intraprocedural maneuver. Some studies discuss the preprocedural score that can predict pancreatitis before the procure. This study discusses a new preprocedural score for post-ERCP pancreatitis.

Thank you for evaluating our manuscript and raising a strength of this manuscript.

(2) Depending on this score that identifies low, moderate, and high-risk patients for post-pancreatitis, so from the start, experienced and well-trained endoscopists can do the procedure or can refer patients to tertiary hospitals or use interventional radiology or endoscopic retrograde cholangiopancreatography.

Thank you for evaluating our manuscript and raising a strength of this manuscript.

(3) The number of patients in this study is sufficient to analyze data correctly.

Thank you for evaluating our manuscript and raising a strength of this manuscript.

Weaknesses:

(1) It is a single-country, retrospective study.

Thank you for this comment. It’s exactly as you said. This is a limitation (Lines 326-327).

(2) Many cases were excluded, so the score cannot be applied to those patients.

Thank you for this valuable comment. The predictive PEP score is not necessary for the excluded patients. The reasons were as follows. Biliary duct cannulation was not attempted in patients for whom it was difficult to identify the Vater papilla. The biliary tract was separated from the pancreas in patients with a past history of choledochojejunostomy, pancreatojejunostomy, or pancreatogastrostomy. PEP risk was thought to be low in these patients and patients who underwent bile duct cannulation via the choledochoduodenal fistula. PEP diagnosis is difficult in patients with acute pancreatitis, whose diagnosis is currently in progress. We added these explanations (Lines 98-106).

(3) Many other studies, e.g., https://link.springer.com/article/10.1007/s00464-021-08491-1, https://pubmed.ncbi.nlm.nih.gov/36344369/, that have been published before discussing the same issue, so what is the new with this score?

Thank you for raising the new reference written by Archibugi et al. in 2023. The novelty of our score is that it is calculated using the factors that are investigated before ERCP procedures. The study written by Archibugi et al. involved procedure time and cannulation attempts for PEP prediction. These two factors are unknown before ERCP procedures. Therefore, a preprocedural predictive risk model for PEP was not created before our study was performed. We added the content of the past study written by Archibugi and included the report as a reference (Lines 65-67, 73-74).

(4) The discussion section needs reformulation to express the study's aim and results.

Thank you for this valuable comment. I have rewritten the first paragraph of the discussion. In the paragraph, we showed that the study achieved the aim on the basis of the results (Lines 245-255).

(5) Why did the authors select these items in their scoring system and did not add more variables?

Thank you for this valuable comment. We selected the items listed in the Japanese guidelines for acute pancreatitis and post-ERCP pancreatitis. We added this description (Lines 123-126). The original references of the guidelines were cited in the first draft version.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Comment1. Please revise these documents: copyright, disclaimer, ethics approval, consent to participate, consent for publication, data and material availability, competing interests, funding, authors' contributions, and acknowledgments.

First, thank you for reviewing our manuscript. We have already described the required information in the “author information” section. The sentences containing this information were proofread in English.

Reviewer #2 (Recommendations for the authors):

Comment 1. It would be best if you did this study in a Prospective way for more validation.

First, thank you for reviewing our manuscript. We have revised our manuscript according to your comments. It’s exactly as you said. These points are limitations (Lines 312-318, lines 326-327). We hope that future validation studies over wider geographic regions will prove our opinions.

Comment 2. The model name should be Acronyum (the first letter of the five items in the risk model).

Thank you for this valuable comment. Sorry, we could not create a memorable model name using the first letter of the five items.

Comment 3. You say that you include the pre-procedure criteria that predict PEP. You state one of the items, pancreatic duct procedure. Do you mean it is a history?

Thank you for this valuable comment. This means that the main purpose is the pancreatic duct. Therefore, the pancreatic duct procedure is listed as “planned pancreatic duct procedures” in Figure 2 (Lines 40-41, 231-234). When an unintended pancreatic duct procedure is performed, we can calculate the risk score by adding two points for “planned pancreatic duct procedures” (Lines 48-49, 247-250).

Comment 4. Regarding calcification, do you mean chronic pancreatitis? It needs more clarification regarding its degree.

Thank you for this valuable comment. We regard pancreatic calcification as a finding of chronic pancreatitis. Pancreatic calcification was defined as the degree that was confirmed by imaging, such as CT, MRI, and EUS. These definitions have been written in the first draft version (Lines 134-137).

Comment 5. Why don't you include young age in the model? Your result found that age less than 50 is significantly associated with PEP.

Thank you for this valuable comment. We selected the PEP risk factors listed in the Japanese guidelines for acute pancreatitis and post-ERCP pancreatitis. Age less than 50 years was listed as a PEP risk factor in the Japanese guidelines for acute pancreatitis. We added this description (Lines 123-126).

Comment 6. There is an ancient reference, some of them in 1994,1996.

Sorry for the old references. These references were written by Cotton et al. 1991, Freeman et al. 1996, and Loperfido et al. 1998. These are still important today. The diagnostic criteria for PEP were determined in the report written by Cotton et al., which is Cotton’s criteria. The other two references are representative reports that described risk factors for PEP, and these two reports were cited in the Japanese guidelines for pancreatitis written by Takada et al. 2022 (Lines 123-126).

Comment 7. In the introduction, you say that the first score includes one of the items for PEP pain during the procedure. It is a little bit strange.

Thank you for this comment. The first PEP risk score did not involve PEP pain but involved pain during the procedure (Line 68).

Comment 8. We know that once ERCP is indicated, you justify the importance of the risk model, stating that if one or more risks are found, we can do EUS or PTD. It is not reasonable to abort the procedure in case of frequent pancreatic duct cannulation or cancel ERCP if pt has one or more risk factors.

Thank you for this valuable comment. If ERCP is performed for high-risk patients, prophylaxes for PEP, such as procedures by experts, pancreatic stent placement, and NSAID suppository insertion, should be performed as much as possible (Lines 281-287, 308-311).

Comment 9. Regarding ERCP pancreatitis criteria, does it include amylase 3t or lipase?

Thank you for this comment. We used Cotton’s criteria for diagnosing PEP. Cotton’s criteria include hyperamylasemia (more than three times the normal upper limit) at least 24 hours after ERCP (114-116).

Comment 10. It is well known that pr with functional biliary disorder has a high incidence of PEP; it doesn't need a manometer for diagnosis. It needs to be included.

Thank you for this comment. Moreover, functional biliary disorders are difficult to diagnose before ERCP procedures (Lines 259-262). The factor that is not apparent before ERCP could not be included in the predictive PEP scoring system.

Comment 11: What is gabexare and nafamost.

Thank you for this comment, and sorry for our insufficient explanation. These compounds include gabexate masilate and nafamostat masilate, which are protease inhibitors. In some institutions, protease inhibitors are used as prophylaxis for PEP. We added “protease inhibitors” (Lines 138-139, Tables 1 and 2).

Reviewer #3 (Recommendations for the authors):

Comment 1. The sample size needs clarification.

First, thank you for reviewing our manuscript. The sample size has been included in the “Methods” section (Lines 157-165).

Comment 2. They need to be mentioned cause they depend on old references in discussion and background.

Thank you for this comment. The previous references were written by Cotton et al. 1991, Freeman et al. 1996, and Loperfido et al. 1998. These are still important today. The diagnostic criteria for PEP were determined in the report written by Cotton et al., which is Cotton’s criteria. The other two references are representative reports that described risk factors for PEP, and these two reports were cited in the Japanese guidelines for pancreatitis written by Takada et al. 2022 (Lines 122-126). In the background and discussion, we added new recent references and information related to the references (Lines 65-67, 285-287, 291-295, 308-311).

Comment 3. Case definition should be added to the methodology.

Thank you for this comment. We added patient information. Please refer to the response against the eLife assessment, weakness, (2).

Comment 4. Do you include all who met the inclusion criteria, or was there any random sampling technique?

No, we did not use random sampling techniques.

Comment 5. What is the value of comparing the development and validation groups? I do not think it adds anything new as if you want to exclude confounders. Has the comparison revealed that a confounder does exist? What was your point of view concerning that?

Thank you for this valuable comment, and sorry for the insufficient explanation. The differences between the development cohort and the validation cohort are important because the goodness of fit for the score could be confirmed in significantly different groups. We added this explanation (Lines 197-199, 251-253).

Version 2

Abstract

Abstract

Background

Post–endoscopic retrograde cholangiopancreatography (ERCP) pancreatitis (PEP) is a severe and deadly adverse event following ERCP. The ideal method for predicting PEP risk before ERCP has yet to be identified. We aimed to establish a simple PEP risk score model (SuPER model: Support for PEP Reduction) that can be applied before ERCP.

Methods

This multicenter study enrolled 2074 patients who underwent ERCP. Among them, 1037 patients each were randomly assigned to the development and validation cohorts. In the development cohort, the risk score model for predicting PEP was established via logistic regression analysis. In the validation cohort, the performance of the model was assessed.

Results

In the development cohort, five PEP risk factors that could be identified before ERCP were extracted and assigned weights according to their respective regression coefficients: -2 points for pancreatic calcification, 1 point for female sex, and 2 points for intraductal papillary mucinous neoplasm, a native papilla of Vater, or the pancreatic duct procedures (treated as “planned pancreatic duct procedures” for calculating the score before ERCP). The PEP occurrence rate was 0% among low-risk patients (≤ 0 points), 5.5% among moderate-risk patients (1 to 3 points), and 20.2% among high-risk patients (4 to 7 points). In the validation cohort, the C statistic of the risk score model was 0.71 (95% CI 0.64–0.78), which was considered acceptable. The PEP risk classification (low, moderate, and high) was a significant predictive factor for PEP that was independent of intraprocedural PEP risk factors (precut sphincterotomy and inadvertent pancreatic duct cannulation) (OR 4.2, 95% CI 2.8–6.3; P < 0.01).

Conclusions

The PEP risk score allows an estimation of the risk of PEP prior to ERCP, regardless of whether the patient has undergone pancreatic duct procedures. This simple risk model, consisting of only five items, may aid in predicting and explaining the risk of PEP before ERCP and in preventing PEP by allowing selection of the appropriate expert endoscopist and useful PEP prophylaxes.

Introduction

Endoscopic retrograde cholangiopancreatography (ERCP) is widely performed as an important diagnostic and therapeutic procedure for pancreaticobiliary diseases. Among endoscopic procedures, ERCP-related procedures are relatively risky. The high-risk adverse events of ERCP include duodenal perforation and bleeding after endoscopic sphincterotomy (EST) and post-ERCP pancreatitis (PEP). The rate of PEP occurrence is 3.1–13.0% (Andriulli et al., 2007, Freeman et al., 1996, Glomsaker et al., 2013, Katsinelos et al., 2014, Kochar et al., 2015, Loperfido et al., 1998). PEP can even become life-threatening. The fatality rate of PEP is 0.1–0.7% (Andriulli et al., 2007, Kochar et al., 2015). Therefore, the decision to perform ERCP should be made carefully, considering each patient’s risk factors for PEP.

To predict an individual patient’s PEP risk, six scoring systems have been devised (Archibugi et al., 2023, Chiba et al., 2021, DiMagno et al., 2013, Friedland et al., 2002, Fujita et al., 2021, Zheng et al., 2020). The first risk scoring system for PEP occurrence was established in 2002. In that study, pain during the procedure, pancreatic duct cannulation, a history of PEP, and the number of cannulation attempts were identified as risk factors for PEP. After the first scoring system was reported, each new scoring system used risk factors that were extracted via multivariate analyses. These included various patient characteristics before ERCP and postprocedural risk factors.

Postprocedural risk factors, such as precut sphincterotomy, procedure time, and difficult cannulation, have been proposed, but it is difficult to predict these risk factors and determine the PEP risk before ERCP. Thus, a new prediction scoring system for PEP before ERCP is desirable. If the risk of PEP can be predicted before ERCP, then the expert endoscopist can perform ERCP from the start, and high-PEP-risk procedures (for example, precut sphincterotomy, multiple cannulation attempts, and inadvertent pancreatic duct cannulation) can be avoided (Testoni et al., 2010, Wang et al., 2009). If biliary cannulation without the use of at least one high-PEP-risk procedure is difficult, other treatments (for example, percutaneous transhepatic biliary drainage (PTBD) or endoscopic ultrasound (EUS)-guided biliary drainage (EUS-BD)) could be considered.

Therefore, we aimed to establish a PEP prediction model using only risk factors that can be gathered before ERCP. Our model was developed and validated with multicenter data from Japan.

Methods

We performed a multicenter retrospective study at six institutions in Japan. This study was approved by the institutional review board of Fukushima Medical University and that of each partner medical institution. All patients agreed to undergo ERCP after providing written consent.

Patients

Among 2,176 patients who underwent ERCP between November 2020 and October 2022, 2,074 were included in this study. The other 102 patients were excluded for the following reasons: past history of choledochojejunostomy, acute pancreatitis, choledochoduodenal fistula, difficulty finding the Vater papilla, past history of pancreatojejunostomy, or past history of pancreatogastrostomy (Figure 1). The predictive PEP score is not necessary for the excluded patients. The reasons were as follows. Biliary duct cannulation was not attempted in patients for whom it was difficult to identify the Vater papilla. The biliary tract was separated from the pancreas in patients with a past history of choledochojejunostomy, pancreatojejunostomy, or pancreatogastrostomy. PEP risk was thought to be clearly low in these patients and patients who underwent bile duct cannulation via the choledochoduodenal fistula. PEP diagnosis is difficult in patients with acute pancreatitis, whose diagnosis is currently in progress.

Flowchart of the inclusion criteria. ERCP, endoscopic retrograde cholangiopancreatography.

Study design

We randomly sampled 50% of the patients as the development cohort and 50% as the validation cohort (Figure 1). In the development cohort, we established a risk scoring system for predicting PEP before ERCP, which was named the support for PEP reduction model (SuPER model). The validation cohort was used to confirm the effectiveness of the scoring system. PEP diagnosis and severity were assessed according to Cotton’s criteria (Cotton et al., 1991). Patients who experienced abdominal pain and had hyperamylasemia (more than three times the normal upper limit) at least 24 hours after ERCP were diagnosed with PEP. Mild PEP was defined as pancreatitis that required prolongation of the planned hospitalization by 2–3 days. Moderate PEP was defined as pancreatitis that required 4–10 days of hospitalization. Severe PEP was defined as pancreatitis that required more than 10 days of hospitalization or intervention or hemorrhagic pancreatitis, phlegmon, or pseudocysts.

To establish the risk score, the risk factors for PEP were investigated via data from the development cohort. To determine the PEP risk score, factors that might be associated with PEP occurrence were investigated. To predict the PEP risk score before ERCP, factors related to patient characteristics and previously scheduled procedures, as reported in the Japanese guidelines for acute pancreatitis and PEP, were selected (Itoi et al., 2024, Takada et al., 2022). The patients’ risk factors included age < 50 years, female sex, a past history of pancreatitis, a past history of PEP, a past history of gastrectomy, pancreatic cancer, intraductal papillary mucinous neoplasm (IPMN), a native papilla of Vater, absence of chronic pancreatitis (CP), normal serum bilirubin (≤ 1.2 mg/dl), and periampullary diverticulum (Ding et al., 2015, Freeman et al., 2001, Freeman et al., 1996, Fujita et al., 2022, Fujita et al., 2021, Masci et al., 2003, Wang et al., 2009, Williams et al., 2007, Zheng et al., 2020). Pancreatic divisum was excluded from the patient risk factor list because pancreatic divisum was observed in only two patients. Pancreatic calcification and a diameter of the main pancreatic duct > 3 mm were considered to indicate CP (Beyer et al., 2023, Sarner and Cotton, 1984).

These imaging findings were confirmed by CT, MRI, and EUS before ERCP. The CT and MRI findings were reviewed by radiologists. IPMNs were diagnosed according to the results of CT, MRI, and EUS. As pre-ERCP prophylaxes for PEP, protease inhibitors (gabexate mesilate or nafamostat mesilate), intravenous hydration, and NSAID suppositories have been used (Fujita et al., 2022). As planned procedure-related risk factors, EST, endoscopic papillary balloon dilation (EPBD), endoscopic papillary large balloon dilation (EPLBD) using a ≥ 12 mm balloon catheter (Itoi et al., 2018), biliary stone removal, ampullectomy, biliary stent material (plastic stent, self-expandable metallic stent (SEMS), or covered SEMS (CSEMS)), inside stent placement, and procedures on the pancreatic duct were evaluated (Freeman et al., 2001, Freeman et al., 1996, Harewood et al., 2005, Kato et al., 2022, Masci et al., 2003, Masci et al., 2001, Testoni et al., 2010, Williams et al., 2007). A biliary stent above the Vater papilla was also assessed as a prophylactic measure against PEP (Ishiwatari et al., 2013).

To demonstrate the independence of the established risk classification, the relationships between it and intraprocedural PEP risk factors (including precut sphincterotomy and inadvertent pancreatic duct cannulation) (Testoni et al., 2010, Wang et al., 2009) were investigated. Because of the retrospective nature of the data, the exact number of cannulations and the number of cannulation attempts were not available. Therefore, multiple cannulation attempts and a prolonged cannulation time could not be investigated as intraprocedural PEP risk factors.

Sample size

The primary aim of this study was to establish a PEP prediction model that could be used to calculate a risk score before ERCP. To construct a prediction model via logistic regression analysis, 10 events per explanatory variable were needed (Wynants et al., 2015). Seven variables were evaluated in the development cohort, so 70 PEP patients were included. Five variables were evaluated in the validation cohort, so 50 PEP patients were necessary. According to a previous systematic review, the rate of PEP occurrence was 9.7% (Kochar et al., 2015). Therefore, at least 722 and 521 patients were included in the development and validation cohorts, respectively.

Statistical analysis

In the development cohort, univariate and multivariate logistic regression analyses were performed to identify the risk factors for PEP. The factors that had a p value < 0.10 in the univariate analysis were included in the multivariate analysis. To construct the scoring system for PEP risk, the factors with p < 0.10 in the multivariate analysis were ultimately included in the risk score model. The factors selected in the multivariate analysis were assigned points according to the regression coefficient (each variable’s risk points = the ratio of the variable’s regression coefficient/minimum regression coefficient). The sum of the assigned points was calculated for each patient, and the patients were classified into three groups (low risk, moderate risk, and high risk) according to the expected rate of PEP occurrence (Friedland et al., 2002). The risk classification system (SuPER model) was also applied to the validation cohort.

With respect to both the development and validation cohorts, the effectiveness of the risk score model was evaluated as follows. The correlations between the risk score, risk classification and PEP occurrence were evaluated via the Cochran–Armitage trend test. The predictive accuracy of the risk score was assessed via the C statistic. The goodness of fit of the model was evaluated via the Hosmer_Lemeshow test. The independence of the established risk classification from the unexpected intraprocedural PEP risk factors was assessed via multivariate logistic regression analyses.

Patients with missing data for variables selected in the risk score model were removed from the final cohort.

Statistical analyses were performed via EZR version 1.62 (Saitama Medical Centre, Jichi Medical University, Saitama, Japan) and SPSS version 26.0 (IBM Corp., Armonk, NY, USA). A p value < 0.05 indicated statistical significance.

Results

Patient characteristics and ERCP outcomes in each cohort

The patient characteristics and ERCP outcomes of each cohort are shown in Table 1. A total of 1037 patients were assigned to each of the development and validation cohorts, including 70 (6.8%) and 64 (6.2%) patients diagnosed with PEP, respectively. Age, history of pancreatitis, and procedures involving the pancreatic duct were significantly different between the development cohort and the validation cohort. The pre-ERCP prophylactic measures used at each hospital differed, and not all patients received prophylaxis.

Construction of the PEP risk scoring system

According to the univariate analyses, age < 50 years, female sex, IPMN, a native papilla of Vater, pancreatic calcification, EST, and procedures on the pancreatic duct had p values < 0.10 (Table 2). According to the multivariate analysis, female sex, IPMN, a native papilla of Vater, pancreatic calcification, and procedures on the pancreatic duct had p values < 0.10. These factors were assigned risk points according to their respective regression coefficients.

The risk score of each patient was calculated as the total number of risk points and ranged from -2 to 7 points (Table 3). The risk score was found to be correlated with PEP occurrence (p < 0.01, Cochran–Armitage trend test). The patients were classified as low (≤ 0 points), moderate (1–3 points), or high risk (4–7 points) for PEP according to the risk score. The PEP rates were 0% (0/327) among the low-risk patients, 5.5% (27/492) among the moderate-risk patients, and 20.2% (39/193) among the high-risk patients. The risk classification was correlated with PEP occurrence (p < 0.01, Cochran–Armitage trend test).

The C statistic of the risk score model was sufficiently high at 0.77 (95% confidence interval (CI) 0.72–0.82) (Table 4). The goodness of fit of the risk score model was also confirmed by the Hosmer–Lemeshow test (p = 0.59).

Validation of the PEP risk scoring system

The risk score was associated with PEP occurrence in the validation cohort (p < 0.01, Cochran–Armitage trend test) (Table 3). We found that 2.4% (8/331) of the patients at low risk, 5.3% (27/513) of those at moderate risk, and 18.0% (29/161) of those at high risk experienced PEP. The risk classification was also correlated with PEP occurrence in the validation cohort (p < 0.01, Cochran–Armitage trend test).

The C statistic of the risk score was 0.71, which was also high in the validation cohort (Table 4). The PEP risk score model showed good fitness according to the Hosmer–Lemeshow test (p = 0.40). According to the above results, the preprocedural PEP risk can be calculated, as shown in Figure 2. The pancreatic duct procedure was treated as a “planned pancreatic duct procedure”. Otherwise, we can calculate both the PEP risk score, which involves the pancreatic duct procedure, and the score, which does not involve the pancreatic duct procedure.

Example of the preprocedural PEP risk checklist. ERCP, endoscopic retrograde cholangiopancreatography. IPMN, intraductal papillary mucinous neoplasm; PEP, post-ERCP pancreatitis.

Risk classification and unexpected PEP risk factors

The relationships between the established risk classification and intraprocedural PEP risk factors are shown in Appendix 1—Table 1. For all patients in the development cohort and the validation cohort, the risk classification was significantly associated with the occurrence of PEP. On the other hand, precut sphincterotomy and inadvertent pancreatic duct cannulation were not significantly associated with the occurrence of PEP.

Discussion

In this multicenter study, we aimed to establish a preprocedural PEP prediction model. As a result, we created a risk scoring system (the SuPER model) using five items that could be measured before performing ERCP. The term “pancreatic duct procedures” was “planned pancreatic duct procedures” in the model. When unintended pancreatic duct procedures are considered, we can calculate both the PEP risk score with pancreatic duct procedures and that without pancreatic duct procedures. With this score, PEP occurrence can be accurately predicted to some degree. Although the development cohort and validation cohort had significantly different patient characteristics and ERCP outcomes, the risk score was good in both cohorts. In addition, the established PEP risk classification was associated with PEP occurrence independent of unpredictable intraprocedural PEP risk procedures.

This risk scoring and classification of PEP has several advantages. First, the score is calculated using only five items, all of which can be easily assessed via medical interviews and imaging (for example, CT). One scoring system included sphincter of Oddi dysfunction (SOD) as a test item (DiMagno et al., 2013). The diagnosis of SOD requires sphincter of Oddi manometry and fulfillment of the criteria for biliary pain, but sphincter of Oddi manometry is not widely used (Cotton et al., 2016). The diagnostic criterion for biliary pain included 8 items, and that for SOD included 15 items. Among the items of the SuPER risk scoring system, pancreatic calcification was assigned -2 points. Its low weighting could be explained by the following. The international conceptual model of CP can be divided into four stages: acute pancreatitis–recurrent acute pancreatitis, early CP, established CP, and end-stage CP (Whitcomb et al., 2016). Established CP patients have already passed the acute pancreatitis–recurrent acute pancreatitis course, and pancreatic calcification has been reported in established CP patients. Acinar dysfunction has also been observed in these patients (Whitcomb et al., 2016). Therefore, patients with pancreatic calcification may have a lower incidence of PEP.

Second, the SuPER risk score can be determined before the ERCP procedure, as the established risk classification was found to be the sole significant factor predicting the occurrence of PEP independent of intraprocedural PEP risk factors. As described in the Background section, precut sphincterotomy, multiple cannulation attempts, and a cannulation time greater than 10 minutes were identified as high-risk factors that cannot be accounted for prior to ERCP (Testoni et al., 2010, Wang et al., 2009). Although the established PEP risk classification was independent of the included intraprocedural risk factors (precut sphincterotomy and inadvertent pancreatic duct cannulation), detailed data on the number of cannulation attempts and the cannulation time were not available. Therefore, to avoid intraoperative procedures associated with a high risk of PEP occurrence, an expert endoscopist can initially perform ERCP for high-PEP-risk patients. In addition, PEP prophylaxis can be administered beforehand for high-PEP-risk patients. As effective prophylaxes for PEP, rectal NSAIDS and pancreatic stent placement have been reported (Akshintala et al., 2021, Choi et al., 2023, Elmunzer et al., 2008, Koshitani et al., 2022, Murray et al., 2003, Sperna Weiland et al., 2022, Sugimoto et al., 2019, Yang et al., 2020). In this report, rectal NSAID use was not identified as a significant factor preventing PEP. One reason for this is that in past reports describing the use of rectal NSAIDs to prevent PEP, patients at high risk for PEP were often treated. In contrast, this study included all patients who underwent ERCP. Another reason might be the difference in dose. One hundred milligrams of rectal diclofenac have been used in most past reports, whereas the efficacy of low-dose rectal diclofenac (25–50 mg) for preventing PEP is under discussion (Maeda et al., 2021, Okuno et al., 2018, Otsuka et al., 2012, Sakai et al., 2023, Takaori et al., 2021, Tomoda et al., 2021). In this study, 12.5–50 mg of diclofenac was used. In Japan, the approved diclofenac dose covered by insurance is 50 mg or less, with the dose typically being lower for elderly patients. Therefore, diclofenac doses ranging from 12.5–50 mg were prescribed by the doctors depending on the age and size of the patients. Pancreatic stent placement itself is one of the procedures performed on the pancreatic duct and is a higher-risk procedure for PEP than endoscopic biliary procedures without an approach to the pancreatic duct (Appendix 2—Table 2). Moreover, pancreatic stent placement has become a prophylactic treatment for PEP in patients who have undergone pancreatography or wire placement to the pancreatic duct (Mazaki et al., 2014, Sugimoto et al., 2019). As described above, pancreatic stent placement was performed along with high-risk-PEP procedures (i.e., guidewire placement to the pancreatic duct or pancreatography); therefore, pancreatic stent placement was grouped with the other endoscopic retrograde pancreatography procedures as “procedures on the pancreatic duct”. In a recent randomized trial involving 1950 patients, the combination of rectal NSAIDs and prophylactic pancreatic stents was more effective for preventing PEP than NSAIDs alone (Elmunzer et al., 2024). For high-PEP-risk patients with high scores, the combination of prophylactic methods might be desirable.

This study has several limitations. First, the study was retrospective, and there were missing data. However, the results reported are trustworthy. The percentage of patients who did not meet the inclusion criteria was not greater than 5%, and the percentage of missing data was not greater than 1%. As described in the Materials and Methods section, patients with missing data for the variables selected in the risk score model were removed from the final cohort. The reliability of the SuPER risk score model was also statistically confirmed. Second, some factors cannot be assessed before ERCP. Additional procedures can be conducted during ERCP, and unplanned pancreatography is often performed in patients who are scheduled for endoscopic cholangiography or biliary treatment. However, the established PEP risk classification was independent of the included intraprocedural risk factors. A planned procedure for accessing the pancreatic duct is listed in the SuPER risk model. Therefore, we can predict the SuPER risk score and classification of patients regardless of whether they have undergone pancreatic duct procedures. Third, this study was performed in a single country. In the future, prospective validation studies over wider geographic regions are needed.

In conclusion, a simple and useful PEP scoring system (SuPER model) with only five clinical items was developed in this multicenter study. This scoring system may aid in predicting and explaining PEP risk and in selecting appropriate prophylaxes for PEP and endoscopic pancreatobiliary procedures for each patient.